```{r}
library(tidyverse)
library(mosaic)
library(dplyr)
library(survey)
```

#read the data into R
```{r}
LearnNewFun
```
```{r}
LearnNewFun0 <- LearnNewFun %>%
  mutate(sexs=ifelse(sexs==1,"Male","Female"))
LearnNewFun0
```
#filter data by province of Alberta and remove invalid responses to the statement of interest
```{r}
LearnNewFun0 %>% 
  filter(al_g02<6, dprov == 48)->LearnNewFun0
LearnNewFun0
```
```{r}
LearnNewFun0 %>% count(sexs)
```
1. Suppose that the target population is province of Alberta. 
#what is the mean and SD of responses regarding learning new things is fun
```{r}
LearnNewFun0 %>% summarize(m=mean(al_g02), sd=sd(al_g02))
```

#what is the mean and SD of responses regarding learning new things is fun for male and for females
```{r}
LearnNewFun0 %>% group_by(sexs) %>%
  summarize(m=mean(al_g02), sd=sd(al_g02))
```
2. 
#divide the population into two sub-populations by males and females
```{r}
Female_Population <- LearnNewFun0[LearnNewFun0$sexs == "Female",]
Male_Population <- LearnNewFun0[LearnNewFun0$sexs == "Male",]
```

#set your seed number
```{r}
set.seed(626)
```

#take a random sample of 100 from the female group
```{r}
Female_Sample <- Female_Population[sample(1:1078, 100),]
Female_Sample$Pop.Size <- 1071
```

#take a random sample of 425 from the male group
```{r}
Male_Sample <- Male_Population[sample(1:1009, 425),]
Male_Sample$Pop.Size <- 1000
```

```{r}
sample_data <- rbind(Female_Sample, Male_Sample)
attach(sample_data)
```

#estimated mean response
```{r}
favstats(al_g02)
```

#corresponding SE
```{r}
library(psych)
describe(al_g02)
```

#estimated mean response for males and for females
```{r}
favstats(al_g02 ~ sexs)
```
#corresponding SE
```{r}
describeBy(al_g02, sexs)
```

3. 
#report the 95% CI for the overall mean response 
```{r}
t.test(al_g02)
```

#difference in mean response between males and females
```{r}
t.test(al_g02 ~ sexs, var.equal = TRUE)
```

4. Give the estimated margin of error for each of the CI that you found 

5. Correct the results by taking into account the population size. 
#finite population correction 
#recall that N = 2087
```{r}
N = 2087
Total.Cases <- N
Total.Cases
```
```{r}
sample_data$Total.Cases <- Total.Cases
Total.Cases
```

#we do fpc since we know the population size 
```{r}
srs.design <- svydesign(data = sample_data, id=~1, fpc=~Total.Cases)
srs.design
```

a. estimated means of overall response and response for males and for females
#estimate the population mean
```{r}
ybar <- svymean(~al_g02, srs.design)
ybar
```

#domain means for each sex
```{r}
svyby(~al_g02, by = ~sexs, design = srs.design, FUN = svymean)
```
b. 95% CI for the overall mean response, and the difference in mean response between males and females
#CI for population 
```{r}
confint(ybar, level = 0.95)
```

#difference in mean response between males and females

c. estimated margin of errors

6. Construct a stratified random sampling to improve your analysis. 
```{r}
detach(sample_data)
attach(LearnNewFun0)
```
#population size by stratum 
```{r}
Nh = table(sexs)
Nh
```
#population size
```{r}
N = nrow(LearnNewFun0)
N
```

#report the allocated sample sizes for each sex group based on the 525 random sample of interest 
```{r}
n = 525
```
```{r}
round((Nh/N)*n)
```

```{r}
library(splitstackshape)
set.seed(626)
str.rs <- data.frame(stratified(LearnNewFun0, "sexs", c(Male = 254, Female = 271)))
attach(str.rs)
```
```{r}
detach(LearnNewFun0)
```
```{r}
Stratified.RS <- str.rs %>% 
  mutate(Pop.Size = ifelse(sexs == "Male", 254, 271),
         Stratum = ifelse(sexs == "Male", 1, 2))
attach(Stratified.RS)
```
```{r}
str.design <- svydesign(id=~1, strata = ~Stratum, fpc=~Pop.Size, data = Stratified.RS)
str.design
```

a. estimated means of overall response and response for males and for females and SE
#estimate the population mean 
```{r}
ybar <- svymean(~al_g02, str.design)
ybar
```

#domain means for each sex
```{r}
svyby(~al_g02, by = ~sexs, design = str.design, FUN = svymean)
```
b. 95% CI for the overall mean response, and the difference in mean response between males and females
#CI for population 
```{r}
confint(ybar, level = 0.95)
```

#difference in mean response between males and females

c. estimated margin of errors